Regression Cube Analysis of Cohort Study Data

### 2015-03-29 draft.tex
added Sylvias revision

The number of features calculated by the CFS algorithm is dependent on the information entropy in the data.
%%
In our epidemiological data we observed a usual number of 10 to 20 features.
%%
For features, such as age or gender, which affect most other features (Recall "Confounder" in Section~\ref{sec:EpidemiologicalData}) the number is significantly larger (about half the features in the whole data set).
%%
The number of features extracted is not directly associated with the overall number, but rather on their information entropy towards each other.
%%
\com{Uli, is this correct this way?}
%%

Now we have a method to derive the interesting regression models in a reasonable time span.
%%
The next section shows ways of abstracting the results to make them visually feasible.

\subsection{Abstracting Regression Results using $R^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/cube}
 \caption{
 (left) Overview visualization using a mosaic plot for the formula $Z \sim X + Y$, where $Z$ assumes feature $Age$.
 %%
 The $R^2$ values extracted from the regression formulas depict the quality-of-fit and are mapped to color saturation.
 %%
 A saturated color shows a strong correlation.
 %%
 (right) Since $Z$ assumes all features $n$ as given by the formula, it yields $n$ mosaic plot visualizations.
 %%
 These represent the slices in our cube.
 %%
 The $R^2$ values of each slice voxel is mapped on opacity in the 3D-view later on, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%% TODO: Cite another paper here!
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded using color on a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the quality-of-fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the $R^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since $R^2$ is scaled between [0,1], it allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the $R^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g. $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
$R^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of $R^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the $R^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
Our goal is to create an overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow}
 \caption{
 Workflow of the analysis using \emph{Regression Cubes}.
 %%
 % The analysis starts with a set of features given by the data set.
 %%
 [\textbf{1.AF}] The workflow starts by declaring formula regarding to specify a hypothesis, or use a predefined formula for a hypothesis-free analysis.
 %%
 [\textbf{2.SI}] The \emph{Regression Cube} is then visualized.
 %%
 The user has the option to either adjust the formula or to derive details on demand on a specific regression.
 %%
 [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or selecting a difference view.
 %%
 The latter is used to compare regression cubes.
 %%
 [\textbf{4.DD}] Details about features using the 2D mosaic plot representation yields insights and hypothesis about feature relations.
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
The analysis workflow follows the Visual Analytics (VA) Mantra of Keim et al. \cite{Keim}:

@@

\textbf{Details on Demand} [\textbf{4.DD}]. Precise information about the individual regression models (coefficients, associated confidence intervals and p-values) can be retrieved based on the data point representatives (e.g. in a hover-modal on a currently selected data point).
\\\\
We use the squared bracket abbreviation for each step to denote the affiliation the system design section later on.
%%
As seen in Fig~\ref{fig:Workflow}, the workflow is highly iterative.
%%
Observations in the 2D-mosaic plot or simply the CFS-based features can trigger new analyses by adjusting the underlying regression formulas.
%%
This can be carried out either to refine the current formula based on observations, or creating a new regression cube for a difference view.
%%
\paragraph{Hypothesis-Free and Hypothesis-Based Analysis.}
%%
Input formulas reflect \emph{hypotheses} about the data.
%%
Using the operators, dynamic variables and dataset features, many different assumptions can be formulated.
%%
To support \emph{hypothesis-free} analysis we provide a default formula:

$Z \sim X + Y$.
%%
This cube represents all possible combinations of two independent features towards all features in the data set, since we do not know which feature(s) are of interest.
%%
Each slice represents a different target feature.
%%
It is therefore suitable for an explorative analysis to give a general impression about relationships in the data set.

Hypotheses about the data are reflected in either how dynamic variables are related using the regression operators.
%%
Also, static features can be added for each regression formula.
%%
Here are a few examples:

%%
$Cancer \sim X + Y + Z$ is an example for the formulation of a hypothesis, where a specific feature is analyzed.
%%
All combinations of three independent features towards the target are analyzed through the cube.

@@

\paragraph{Cube Comparison.}
%%
Regression Cubes can be compared by creating difference views.
%%
One cube (formula) acts as reference.
%%
The absolute difference in $R^2$ values towards the second cube is calculated, yielding a difference cube showing only the differences between the two formulas.
%%
It can for example be utilized for comparing the influence of one feature towards the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$)
%%
% \paragraph{Formula Types.}
% %%
% Input formulas reflect \emph{hypotheses} about the data.
% %%
% Using the operators, dynamic variables and dataset features, many different assumptions can be formulated.
% %%
% Here we show just a few examples on how formulas can serve different purposes.
%
% $Z \sim X + Y$ is the default formula.
% %%
% This cube represents all possible combinations of two independent features towards all features in the data set, since we do not know which feature(s) are of interest.
% %%
% Each slice represents a different target feature.
% %%
% It is therefore suitable for a \emph{hypothesis-free} explorative analysis to give a general impression about relationships in the data set.

>>>>

The number of features calculated by the CFS algorithm is dependent on the information entropy in the data.
%%
In our epidemiological data, we usually observed a number of 10 to 20 features.
%%
% Was heißt der Termin in Klammern? 1. Weg; 2. "ca. 50% of the feature set 
For features, which affect the most other features (Recall "Confounder" in Section~\ref{sec:EpidemiologicalData}), e.g., age or gender,  the number is significantly larger (about half the features in the whole data set).
%%
The number of features extracted is not directly associated with the overall number, but rather on their information entropy towards each other.
%%
\com{Uli, is this correct this way?}
%%

With this method, we are able to derive the interesting regression models in a reasonable time span.
%%
The next section shows ways of abstracting the results to make them visually feasible.

\subsection{Abstracting Regression Results Using $R^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/cube}
 \caption{
 (left) Overview visualization using a mosaic plot for the formula $Z \sim X + Y$, where $Z$ assumes the feature $Age$.
 %%
 The $R^2$ values extracted from the regression formulas depict the quality-of-fit and are mapped to color saturation (a saturated color shows a strong correlation).
 %%
 (right) Now, $Z$ is set to all features $n$ and yields $n$ mosaic plot visualizations.
 %%
 These represent the slices in our cube.
 %%
 The $R^2$ values of each slice voxel is mapped on opacity in the 3D-view later on, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%% TODO: Cite another paper here!
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded with color in a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the quality-of-fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the $R^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since $R^2$ is scaled between [0,1], it allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the $R^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g., $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
$R^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of $R^2$ values derived from different regression cubes (e.g., $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
% e.g. wenn man es wählen kann
% i.e. wenn es so ist
Therefore, the $R^2$ results of different regression methods are encoded using different colors (i.e., blue for linear regression and red for logistic regression).
%%
% Adverb immer vors Hauptverb oder Satzende
Thus, the cube can be easily extended using other regression types.
%%
For cubes with a fixed target feature, e.g., $Cancer \sim X + Y + Z$, no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
Our goal is to create an overview visualization for a data set, but we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
The different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow}
 \caption{
 Different workflow types of the analysis using \emph{Regression Cubes} (based on~\cite{Keim}).
 %%
 % The analysis starts with a set of features given by the data set.
 %%
 [\textbf{1.AF}] The workflow starts by declaring a formula regarding to specify a hypothesis, or use a predefined formula for a hypothesis-free analysis.
 %%
 [\textbf{2.SI}] The \emph{Regression Cube} is then visualized.
 %%
 The user has the option to either adjust the formula or to derive details on demand on a specific regression.
 %%
 [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or selecting a difference view.
 %%
 The latter is used to compare regression cubes.
 %%
 [\textbf{4.DD}] Details about features using the 2D mosaic plot representation yields insights and hypothesis about feature relations.
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
% Übergang ist recht holprig gewesen
Our Regression Cube models is well-suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al. \cite{Keim}:

@@

\textbf{Details on Demand} [\textbf{4.DD}]. Precise information about the individual regression models (coefficients, associated confidence intervals and p-values) can be retrieved based on the data point representatives (e.g. in a hover-modal on a currently selected data point).
\\\\
We use the squared bracket abbreviation for each step to denote the affiliation the system design section later on.
%%
As seen in Fig~\ref{fig:Workflow}, the workflow is highly iterative.
%%
Observations in the 2D mosaic plot or simply the CFS-based features can trigger new analyses by adjusting the underlying regression formulas.
%%
This can be carried out either to refine the current formula based on observations, or creating a new regression cube for a difference view.
%%
\paragraph{Hypothesis-Free and Hypothesis-Based Analysis.}
%%
Input formulas reflect \emph{hypotheses} about the data.
%%
Using the operators, the dynamic variables and the dataset features, many different assumptions can be expressed.
%%
To support \emph{hypothesis-free} analysis, we provide a default formula:

$Z \sim X + Y$.
%%
This cube represents all possible combinations of two independent features towards all features in the data set, since we do not know which features are of interest.
%%
Each slice represents a different target feature.
%%
It is therefore suitable for an explorative analysis to give a general impression about relationships in the data set.

% warum war da ein "either", sollte der Satz etwas anderes ausdrücken?
Hypotheses about the data are easily built up by relating dynamic variables with the regression operators.
%%
Furthermore, static features can be added for each regression formula.
%%
Here are a few examples:

%%
$Cancer \sim X + Y + Z$ is an example for the formulation of a hypothesis, where the specific feature \textit{Cancer} is analyzed.
%%
All combinations of three independent features towards the target are analyzed through the cube.

@@

\paragraph{Cube Comparison.}
%%
Regression Cubes can be compared by creating difference views.
%%
One cube (extracted with one formula) acts as reference.
%%
The absolute difference in $R^2$ values towards the second cube is calculated, yielding a difference cube showing only the differences between the two formulas.
%%
For example, it can be utilized for comparing the influence of a single feature towards the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$)
%%
% \paragraph{Formula Types.}
% %%
% Input formulas reflect \emph{hypotheses} about the data.
% %%
% Using the operators, dynamic variables and dataset features, many different assumptions can be formulated.
% %%
% Here we show just a few examples on how formulas can serve different purposes.
%
% $Z \sim X + Y$ is the default formula.
% %%
% This cube represents all possible combinations of two independent features towards all features in the data set, since we do not know which feature(s) are of interest.
% %%
% Each slice represents a different target feature.
% %%
% It is therefore suitable for a \emph{hypothesis-free} explorative analysis to give a general impression about relationships in the data set.
